\Title

### 2013-02-09 supplementary.tex
Edits in main text and supplement. Supplement needs more work. Methods need to be somewhat more detailed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Template for a revtex article
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{{\color{red}#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini* and Richard~A.~Neher*}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\Title}
\author{\Author}
\affiliation{\textsf{*}\Affiliation}
\date{\today}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p7}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
\caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
changes in the same patient. Panels C and D) the same plots from patient p7, whose
sequences show signs of compartimentalization.}
\label{fig:aftp}
\end{center}
\end{figure}
Not all of the patients studied in refs.~\citealp{shankarappa_consistent_1999} and
\citealp{bunnik_autologous_2008} were analyzed. The reason is that patients p4,
p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and patients ACH19542
and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
compartimentalization (population structure) during infection, which would have
been an irreparable confounder for our analysis. The difference is clear if
PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
This is shown in \figurename~\ref{fig:aftp}: late sequences subdivide in two
well-defined clusters -- this characteristic is absent from all the patient
sequences used in our main analysis. On the right, allele frequency trajectories from the same
patient are plotted: evidently, no derived allele is fixed after 1000 days
from seroconversion, but a lot of mutations swarm to frequencies around 0.5-0.8. Of
course, the fixation probability of new mutations in such a sample is very low,
but this is an effect of population structure rather than any
of the processes discussed in the main text (time-dependent selection, competition
of mutants within the same epitope).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonsynonymous changes outside of variable regions are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{synmut_conservation_4fold_synnonsyn}
\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
{\it gag}. Nonsynonymous changes are, overall, much less common than synonymous
ones.}
\label{fig:synnonsyncons}
\end{center}
\end{figure}
In the main text, it is argued that, whereas the deleterious effect of
synonymous changes is very small, most nonsynonymous mutations outside of the
variable regions are much worse. To show this, we take a multiple sequence alignment of
the {\it gag} gene of subtype-B HIV sequences from the LANL website (filtered
sequences only, version 2011)~\cite{LANL2012} and look at diversity at fourfold degenerate
codons. The ratio of observed versus possible mutations is much lower for
nonsynonymous changes than synonymous ones, as shown in
\figurename~\ref{fig:synnonsyncons} in a cumulative histogram. From this
evidence (which is highly statistically significant) we conclude that, as a
general rule, nonsynonymous changes that are not currently escape mutations
are likely to incur in large fitness costs.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Template for a revtex article
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{{\color{red}#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini* and Richard~A.~Neher*}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\Title}
\author{\Author}
\affiliation{\textsf{*}\Affiliation}
\date{\today}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p7}
\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
\caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
changes in the same patient. Panels C and D) show analogous plots for data from
patient p7. Samples after day 1000 split into two clusters in the PCA and no
mutations that arise after day 1000 fix, presumably because they are restricted
to one subpopulation. }
\label{fig:aftp}
\end{center}
\end{figure}
Patients p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
patients ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
compartimentalization (population structure) during infection. This population
structure is easy to spot after a principle component analysis (PCA) of all
sequences from all samples from one patients. When the projections onto the
first 3 PCs for each sequence, the above mentioned patients show bifurcation
into two clusters at later time points. Since mutations that arise in one subpopulation cannot fix population wide, such population
structure confounds our analysis and these patients were exclude.

Similar, but less obvious, conclusions can be drawn from the allele frequency
trajectories plotted next to the PCA plots: no allele that arose after 1000
days from seroconversion ever fixes, but a most mutations hover at frequencies
around 0.5-0.8. Fluctuations of these allele frequencies probably reflect
changing relative sizes of the subpopulations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonsynonymous changes outside of variable regions are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{synmut_conservation_4fold_synnonsyn}
\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
{\it gag} in the LANL reference panel. Nonsynonymous sites are much
less diverse than 4-fold synonymous sites. \comment{The title is wrong. How
do you decide on 4-fold degeneracy? (consensus amino acid???)}.}
\label{fig:synnonsyncons}
\end{center}
\end{figure}
\comment{this can use streamlining. Why not put is all into the figure caption}
In the main text, it is argued that, whereas the deleterious effect of
synonymous changes is very small, most nonsynonymous mutations outside of the
variable regions are much worse. To show this, we take a multiple sequence alignment of
the {\it gag} gene of subtype-B HIV sequences from the LANL website (filtered
sequences only, version 2011)~\cite{LANL2012} and look at diversity at fourfold degenerate
codons. The ratio of observed versus possible mutations is much lower for
nonsynonymous changes than synonymous ones, as shown in
\figurename~\ref{fig:synnonsyncons} in a cumulative histogram. From this
evidence (which is highly statistically significant) we conclude that, as a
general rule, nonsynonymous changes that are not currently escape mutations
are likely to incur in large fitness costs.
